Is high social class always beneficial for survival? A study of Northern Sweden

### 2017-11-16 article3.Rmd
Language corrections, part 2

Differences in available information in the two datasets as well as within
the Linnaeus data makes it necessary to apply different approaches when it
comes to identification of presence periods. The Poplink data provides us with exact dates or at least year of start and exit of presence, allowing us
to have full and continuous control of the \emph{de jure} population. This is
however not the case with the Linnaeus database. For the period 1976--2001
we use presence in the censuses 1975--1990 and information on deaths from the National
Board of Health and Welfare. The population in the Skellefteå and Umeå regions are those residing there according to the censuses. Each census constitute the baseline for which persons are followed up during the five years until the next census (1975 census have a follow-up from 1 January 1976 until the end of 1980). The expection is the census 1990 where the follow-up is until 2001. Thus, the deaths do not necessarily take place in our region for the Linnaeus analyses, although this is usually the case.

For the last period, 2002--2013, we use the yearly population registers
(LISA data), together with information on deaths from National Board of
Health and Welfare. This allows for full control of the presence periods
in the studied regions.

@@

Mortality differences are analysed according to social class, based on a modified form 
of the classification scheme Hisclass [@lema11]. The only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification
schemes in the two databases, is occupations. Thus, analysing according to social class is apart from
being a scientific choice, also a practical one. Social position is treated as a time-dependent variable during working age but defined from the last occupation in working age for the elderly and retired population, i.e. from age 65 until death or last observation.

The availability of information on occupations varies over time and between
sources. Occupations in Poplink are coded according to the original DDB coding system that then has been adapted to Hisclass. For the Linnaeus data, HISCO is recoded into Hisclass. The censuses include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
Yrkesklassificering). We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our definition of presence periods. For the period 1991--2000 the social class in the census 1990 is considered to be kept until 2001.

The use of different data sets forces us to apply different approaches
towards defining the social position. Poplink usually only provides occupation for the head of household, thus underestimating female working-force participation as well as that of
adult children residing with their parents [@vik10]. For the Poplink period, we have chosen to categorise wives according to position of the head of household, usually
the husband/father, assuming that the family shares the same socio-economic
position.

Female labour is much better covered in the Linnaeus database. A
complication is however that it is  difficult to define households in the
same way as in Poplink. Extra-marital cohabitation became common and female
labour force participation developed as the norm in Sweden during this
period. The occupation of the husband as signifier of social class became,
if not obsolete, at least less relevant. All included persons are signified
by their individual occupation for this period. Still, the results for the
different periods are not completely comparable. Notice however that there was a strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data [@bjork92].

Hisclass, the classification system used as a basis for our categorisation, is a "...
HISCO-based historical international social class scheme" [@lema11; @lema02]. 
The different classes in Hisclass represent distinct
categories, based on whether the work is manual or non–manual, skill level,
supervision and sector. Implicitly it reflects large differences in access to
economic and other resources, status, social power et cetera. Longitudinal analyses of social classes for such a long period covered in this study involves several problems, related to changes in occupational structure and the meaning and contents of occupations and social positions. We have chosen to work with a broad definition of classes, merging the original 12 classes in Hisclass into the following three and an additional fourth category of unknown class:

@@

Figures \@ref(fig:hiscskumm) and \@ref(fig:hiscskumw) show the distribution
(per cent) of exposure time according to class for men and women
respectively. The category of workers is fairly stable over time. The
highest class has increased substantially, although from a very low level
in the 19th century. The increase of academics in connection to the
establishment of Umeå University explains much of this increase. What is
not apparent from Figure \@ref(fig:hiscskumm) is the transfer within the middle
group from a majority of farmers during the 19th century to lower white
collar from the middle of the 20th century. Furthermore we observe a rather high percentage of missing social class in the early 19th century. For men the proportion is somewhat higher in the age group 65 and above, those that were retired, but still low enough to be used in our analyses (results not shown). If we disregard the very first period, there were however very few with missing occupation until 1950.

@@

The proportion with missing values among women is however a larger problem,
see Figure \@ref(fig:hiscskumw). Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way womens' social position is defined. The frequency of missing social class is however quite high during particularly the first two periods with data from the Linnaeus database---from 1976. Since female social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce is that the proportion missing increases. For the later periods,
the results are fairly complete.

@@

Marital status, see Figure \@ref(fig:civstskum), is treated as a time-dependent variable in the data
file. The statuses used are unmarried, married and dissolved marriages
(without distinguishing between widowed and divorces). If no explicit
status is given in the sources and partner is missing, the status has been
set to unmarried. Notice that it has not been possible to identify cohabitation.

@@

In 19th and early 20th century Sweden (as in other countries), mortality
was much higher in urban environments [@hani94; @edv92]. The unsanitary and overcrowded
living conditions as well as the risks for importing infections made the
health risks higher in towns and cities. In the present paper, the variable **urban** controls
for this aspect by distinguishing between urban and rural residence for the periods after 1900, based on parish or place of residence.

## Periods for analysis

We have made separate analyses for the different time periods. The period
1801--1950 is divided into six subperiods of length 25 years each. The 19th century
periods represent a mainly pre-industrial society dominated by agriculture
and with only basic welfare provisions. During the years 1901--1950,
industrialisation started and the urban environments increased in
population size. The modern Swedish welfare state began to develop. For the 
period 1976 onwards, there are separate analyses on 5- or 6-year periods. Sweden had developed into a wealthy welfare society with low income inequality,
particularly until the 1990s. A financial crisis struck Sweden in the early
1990s, which among other things increased unemployment and changed parts of
the welfare system towards being less generous. The start of the
last period (2008--2013) coincides with the international financial
crisis. During the last decades income inequality has increased considerably.

@@

The proportional hazards model in a survival analysis context is used,
allowing for adjustment for marital status and rural/urban environment.
The central explanatory variable, social class, does not follow the
proportionality property, therefore the analyses are stratified with respect to the
variable, and the main results are presented graphically.

Another practical problem is the size of the data set:
It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
 `r n.d + pn.d` deaths.
This makes it very inconvenient to use the nowadays so popular semi-parametric
technique known as
\emph{Cox regression} [@cox72]. Instead we use a parametric model
that allows for reduction of the data
to \emph{sufficient statistics} [@fisher22]. The idea is to keep all
relevant covariates \emph{categorical}, and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three marital statuses, ten periods, two
urban/rural groups, four social classes,
ten age groups, and 22 cohorts, in all 105600 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 14084. The data set is thus reduced in
size by 99 per cent.

The parametric model we use for the baseline hazard function is piece-wise
constant on five-year intervals, so we arrive at the following proportional
hazards regression model:

@@

Since all components of $\mathbf{x}$ are categorical covariates, the
estimation of \@ref(eq:prop) can be performed in two ways, first by the
original data set and a survival analysis routine handling a piecewise
constant hazard, or second, by the reduced data set (table) and
\emph{Poisson regression}. The second way is faster than the first by a
factor of size at least 100, often much more, and
the point is that the two ways give exactly the same results. This is
because the table is a \emph{sufficient statistic} for \@ref(eq:prop) and
the original data set.

The analyses are performed in the **R** environment for statistical computing
and graphics [@cran], especially using the package **eha**
[@gb12;@eha]. In the analysis, data are divided into two age groups,
40--64 and 65--89 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period are analysed separately, 
but we then combine results in order to show the development of the social patterning of mortality during the studied period.

The important explanatory variable, social class, is included in the
proportional hazards models as a *stratification* variable. The main
reason for this choice is that the effect of social class on survival is
non-proportional, that is, it varies with age. With the reduced data set
and Poisson regression, this is simply achieved by introducing an
interaction term between **age** and social class. Then, the main
results are presented graphically.  The comments on the results are based on
the figures and focuses on the central variable in our study, social class.

@@

```{r getske4060}
cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women."
scap1w <-  "Cumulative hazards, ages 40--64, women."
cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men."
scap1m <-  "Cumulative hazards, ages 40--64, men."
##
cap0 <- "Probability of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom)."
scap0 <- "Probability by social class, ages 40--64."
cap12 <- "Probability of dying before age 65 for a 40 year old person standardised by social class and period, women (top) and men (bottom)."
scap12 <- "Probability standardised by social class, ages 40--64."
cap12log <- "Probability of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom). The last decades."
scap12log <- "Probability by social class, ages 40--64, last years."

cap2w <-  "Cumulative hazards for social class by time period, ages 65--89, women."
scap2w <-  "Cumulative hazards, ages 65--89, women."
cap2m <-  "Cumulative hazards for social class by time period, ages 65--89, men."
scap2m <-  "Cumulative hazards, ages 65--89, men."
##
cap22 <- "Probability of dying before age 90 for a 65 year old person by social class and period, women (top) and men (bottom). Standardised."
scap22 <- "Probability by social class, ages 65--89."
cap22log <- "Probability of dying before age 90 for a 65 year old person by HISCLASS and period, women (top) and men (bottom). The last decades."
scap22log <- "Probability by social class, ages 65--89, last years."

@@

From Table \@ref(tab:cumswtab) it is obvious that the statistical significance is strong.
The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**. This indicates that fundamental changes has taken place in the relation between the variables during our study period.

@@

Figure \@ref(fig:cumsw) shows the summed hazards of dying in the age interval
40--64 during the different periods. Obviously, but not surprisingly, mortality
is decreasing over time for both sexes. Women have consistently better survival
than men. This is a well-known phenomenon in Swedish
demographic history [@wil99; @suwi07]. We also observe a
substantial decline in mortality for both sexes, resulting in smaller
absolute differences between social classes. We must however distinguish between differences in absolute risks compared to relative risks [@hoetal16]. The compression that is
visible for later decades is partly explained by the decreasing general
mortality and is not necessarily representing a homogenisation of relative risks.

By standardising the levels in each time period for each class, we 
illustrate the relative risks in Figure \@ref(fig:cumswstand). In the rest of 
the paper, we only present these
relative mortality levels for the different classes since they are the main focus in this study.

@@

The changing economy have some consequences for the analysis, particularly when
it comes to the agricultural sector. While farmers were common occupations
up until the first half of the 20th century, these occupations has constantly
decreased in numbers thereafter. This means that the middle category has
a different character during the last periods.

The most striking result of the analysis is the different patterns for men
and women. The social pattern of mortality among women is quite much in
accordance with previous analyses on other regions in northern Sweden and for other periods [@edli11; @edbr12]. The highest social class usually have comparatively low mortality (except for the first period when however the statistical power was weak), while
working class women have high, but almost constantly about the same level as the middle category. There is however no substantial change in the relative risks between classes except for the periods 1976 onwards when a divide between workers and the middle category appears. A consisten social gradient has developed. The group with unknown social class is distinguished by much higher mortality from the 1976 onwards. This can however have to do with how we have defined social position for women before and after 1950. After 1975 it is based
on their own occupation. Being without occupation probably include women that
are more frail when it comes to health or are vulnerable due to social characteristics.

A completely different pattern appears among men, being the most surprising finding in
our analyses. The groups we expect to have the lowest mortality, if
we assume that access to vital resources and belonging to a high social class
determines survival, instead have by far the highest hazards of dying
during the nineteenth and much of the 20th century. Workers on the
other hand have comparatively low mortality until 1950, at the same
level as the middle class. The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead we find better survival among workers compared to the elite.

The relative positions change radically towards
the end of the 20th century. From the 1980s, there is a clear advantage when
it comes to survival for the elite. The relative
differences are however more compressed compared to the periods until 1950. The
expected social gradient in mortality has developed. This is also what we
usually find in analyses of social health inequalities in present-day
Sweden as well as in most other countries, results that has formed the basis for many of the contemporary epidemiologic and demographic theories. In the regions we are studying,
this is however a quite recent phenomenon. Corresponding to what we find
for women, those without known occupation have much higher mortality than
the others from the late 1970s onwards. The assumption made for women of
unknown social class is probably even more relevant for men. These men probably
represent groups in society that are particularly vulnerable.

A further observation can be made from Figure \@ref(fig:cumsw). The gender
difference in death hazards is extremely large for the elite group from
1851--1950, men having about three times higher risks than women in the
same class (basically living in the same households). In the other classes, the difference was much smaller. It was men in high position that experienced the largest risks.

@@

We now turn to the elderly, those that in most cases have left the workforce.
Corresponding to the previous analysis, we use the value of the cumulative hazards at
duration 25 (age 90), starting from age 65 as a summary measure. From Table \@ref(tab:cumswtabold) it is obvious that the statistical significance still is strong.

@@

The general pattern and development of mortality (not shown here) do not
differ that much from the younger age group, of course representing a much
higher mortality level earlier in history. Mortality among women was, as
expected, lower than among men. We find a clear decline in mortality, but
the fastest decline did not set in definitely until some time after the
Second World War, that is, later than for the age
group 40--64 years. This corresponds to the Swedish development where we
find a turning point at this time with better survival among elderly, first for women and a couple of decades later for men.

Concerning the social pattern, the pattern is less clear compared to what
we find for the younger age group in line with the status levelling hypothesis (see Figure \@ref(fig:cumswold)). For women the results are mixed. We do not find the
same clear advantage for elite women during the period before 1950, but neither
do we find any large disadvantage for working class elderly women at that
time. In most periods, the social differences were small, particularly between the worker and the middle group, and the elite even had somewhat higher mortality during the early 20th century. The social differences continued to be small until the last decades of the
20th century. Only during the two last periods (after the year 2000) is it
possible to establish the expected social gradient in mortality.

For men we find indications of the same intriguing change among elderly as
for the adult male population 40--64 years, but at a much lower level. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
the 20th century. Not until the second half of the 20th
century is the survival better for the highest social class while
workers in the age of retirement had a little higher mortality. The difference between the worker and middle group have in most periods been small.

@@

Previous studies have scrutinised different aspects of the social
pattern of mortality in historical northern Sweden for some time, focusing
on different regions, age groups and time periods [@edv92; @edv04; @edli11; @edbr12]. When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economicdifferences were often but not always confirmed [@edv92; @edv04],
something that has been found in other Swedish studies as well [@bredro02; @bube01; @mol17].

>>>>

Differences in available information in the two datasets as well as in
the Linnaeus data make it necessary to apply different approaches when it
comes to the identification of presence periods. The Poplink data provide us with exact dates, or at least year of start and exit of presence, allowing us
to have full and continuous control over the \emph{de jure} population. This is
not the case with the Linnaeus database, however. For the period 1976--2001
we use presence in the censuses 1975--1990 and information on deaths from the National
Board of Health and Welfare. The populations in the Skellefteå and Umeå regions are those residing there according to the censuses. Each census constitute the baseline for which persons are followed up during the five years until the next census (e.g. the 1975 census has a follow-up from 1 January 1976 until the end of 1980). The expection of thisis the 1990 census for which the follow-up is until 2001. Thus, the deaths do not necessarily take place in our region for the Linnaeus analyses, although this is usually the case.

For the last period, 2002--2013, we use the yearly population registers
(LISA data) together with information on deaths from the National Board of
Health and Welfare. This allows for full control over the presence periods
in the studied regions.

@@

Mortality differences are analysed according to social class, based on a modified form 
of the classification scheme Hisclass [@lema11]. The only, and obvious, alternative for conducting a longitudinal study of socio-economic mortality differences with consistent classification
schemes in the two databases is occupations. Thus, analysing according to social class, apart from
being a scientific choice, is also a practical one. Social position is treated as a time-dependent variable during working age but defined from the last occupation in working age for the elderly and retired population; i.e. from age 65 until death or last observation.

The availability of information on occupations varies over time and between
sources. In Poplink, occupations are coded according to the original DDB coding system that has then been adapted to Hisclass. For the Linnaeus data, HISCO is recoded into Hisclass. The censuses include occupational codes based on the system used in Sweden at the time, NYK (Nordisk
Yrkesklassificering). We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our definition of presence periods. For the period 1991--2000 the social class in the 1990 census is considered to be maintained until 2001.

The use of different data sets forces us to apply different approaches
to defining the social position. Poplink usually only provides the occupation of the head of household, thus underestimating female labour force participation as well as that of
adult children residing with their parents [@vik10]. For the Poplink period, we have chosen to categorize wives according to the position of the head of household, usually
the husband/father, assuming that the family shares the same socio-economic
position.

Female labour is much better covered in the Linnaeus database. A
complication, however, is that it is  difficult to define households in the
same way as in Poplink. Extramarital cohabitation became common and female
labour force participation developed as the norm, in Sweden during this
period. The husband's occupation, as signifier of social class became,
if not obsolete, at least less relevant. All included persons are signified
by their individual occupation for this period. Still, the results for the
different periods are not completely comparable. Noti, however, that there was strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way with the use of occupations on individual level instead of household level for defining social class based on the Linneaus data [@bjork92].

Hisclass, the classification system used as a basis for our categorization, is a "...
HISCO-based historical international social class scheme" [@lema11; @lema02]. 
The different classes in Hisclass represent distinct
categories, based on whether the work is manual or non–manual, skill level,
supervision, and sector. Implicitly it reflects large differences in access to
economic and other resources, status, social power, etc. Longitudinal analyses of social classes for such a long period covered in this study present several problems, related to changes in occupational structure and the meaning and content of occupations and social positions. We have chosen to work with a broad definition of classes, merging the original 12 classes in Hisclass into the following three and an additional fourth category of unknown class:

@@

Figures \@ref(fig:hiscskumm) and \@ref(fig:hiscskumw) show the distribution
(per cent) of exposure time according to class for men and women,
respectively. The category of workers is fairly stable over time. The
highest class has increased substantially, albeit from a very low level
in the 19th century. The increase in academics in connection to the
establishment of Umeå University explains much of this increase. What is
not apparent in Figure \@ref(fig:hiscskumm) is the transfer within the middle
group from a majority of farmers during the 19th century to lower white-collar from the middle of the 20th century. Furthermore we observe a rather high percentage of missing social class in the early 19th century. For men the proportion is somewhat higher in the age group 65 and above, those who were retired, but still low enough to be used in our analyses (results not shown). If we disregard the very first period, however, there were very few with occupation missing until 1950.

@@

The proportion with missing values among women is a larger problem, however;
see Figure \@ref(fig:hiscskumw). Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way women's social position is defined. However, the frequency of missing social class is quite high during particularly the first two periods of data from the Linnaeus database---from 1976. Since female social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce is that the proportion of missing increases. For the later periods,
the results are fairly complete.

@@

Marital status (see Figure \@ref(fig:civstskum)), is treated as a time-dependent variable in the data
file. The statuses used are unmarried, married and dissolved marriage
(without distinguishing between widowed and divorced). If no explicit
status is given in the sources and the partner is missing, the status has been
set to unmarried. Note that it has not been possible to identify cohabitation.

@@

In 19th and early 20th century Sweden (as in other countries), mortality
was much higher in urban environments [@hani94; @edv92]. The unsanitary and overcrowded
living conditions, as well as the risk of importing infections, made the
health risks higher in towns and cities. In the present paper, the variable **urban** controls
for this aspect by distinguishing between urban and rural residence for the periods after 1900, based on parish or place of residence.

## Periods of analysis

We have done separate analyses for the different time periods. The period
1801--1950 is divided into six subperiods of 25 years each. The 19th century
periods represent a mainly pre-industrial society dominated by agriculture
and with only basic welfare provisions. During the years 1901--1950,
industrialization started and the urban environments increased in
population size. The modern Swedish welfare state began to develop. For the 
period 1976 onwards, there are separate analyses of five- or six-year periods. Sweden had developed into a wealthy welfare society with low income inequality. Then in the early 1990s, a financial crisis struck Sweden, which among other things increased unemployment and changed parts of
the welfare system towards being less generous. The start of the
last period (2008--2013) coincides with the international financial
crisis. In the most recent decades income inequality has increased considerably.

@@

The proportional hazards model in a survival analysis context is used,
allowing for adjustment for marital status and rural/urban environment.
As the central explanatory variable, social class, does not follow the
proportionality property, the analyses are stratified with respect to the
variable, and the main results are presented graphically.

Another practical problem is the size of the data set:
It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
 `r n.d + pn.d` deaths. This makes it highly inconvenient to use the currently so popular semi-parametric technique known as \emph{Cox regression} [@cox72]. Instead we use a parametric model
that allows for a reduction of the data to \emph{sufficient statistics} [@fisher22]. The idea is to keep all relevant covariates as \emph{categorical} and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three marital statuses, ten periods, two
urban/rural groups, four social classes,
ten age groups, and 22 cohorts; in all, 105,600 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 14,084. The data set is thus reduced in
size by 99 per cent.

The parametric model we use for the baseline hazard function is piecewise
constant at five-year intervals, so we arrive at the following proportional
hazards regression model:

@@

Since all components of $\mathbf{x}$ are categorical covariates, the
estimation of \@ref(eq:prop) can be performed in two ways; based on the
original data set and a survival analysis routine handling a piecewise
constant hazard;  or based on the reduced data set (table) and
\emph{Poisson regression}. The second way is faster than the first by a
size factor of at least 100, often much more, and
the point is that the two ways give exactly the same results. This is
because the table is a \emph{sufficient statistic} for \@ref(eq:prop) and
the original data set.

The analyses are performed in the **R** environment for statistical computing
and graphics [@cran], especially using the package **eha**
[@gb12;@eha]. In the analysis, data are divided into two age groups,
40--64 and 65--89 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period is analysed separately, 
but we then combine results in order to show the development of the social patterning of mortality during the studied period.

The important explanatory variable, social class, is included in the
proportional hazards models as a *stratification* variable. The main
reason for this choice is that the effect of social class on survival is
non-proportional; that is, it varies with age. With the reduced data set
and Poisson regression, this is simply achieved by introducing an
interaction term between **age** and social class. Then, the main
results are presented graphically.  The comments on the results are based on
the figures and focus on the central variable in our study, social class.

@@

```{r getske4060}
cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women."
scap1w <-  "Cumulative hazards, ages 40--64, women."
cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men."
scap1m <-  "Cumulative hazards, ages 40--64, men."
##
cap0 <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
scap0 <- "Probability by social class, ages 40--64."
cap12 <- "Probability of dying before age 65 for a 40-year-old standardized by social class and period, women (top) and men (bottom)."
scap12 <- "Probability standardized by social class, ages 40--64."
cap12log <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom). The last decades."
scap12log <- "Probability by social class, ages 40--64, last years."

cap2w <-  "Cumulative hazards for social class by time period, ages 65--89, women."
scap2w <-  "Cumulative hazards, ages 65--89, women."
cap2m <-  "Cumulative hazards for social class by time period, ages 65--89, men."
scap2m <-  "Cumulative hazards, ages 65--89, men."
##
cap22 <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). Standardised."
scap22 <- "Probability by social class, ages 65--89."
cap22log <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). The last decades."
scap22log <- "Probability by social class, ages 65--89, last years."

@@

Looking at Table \@ref(tab:cumswtab) it is obvious that the statistical significance is strong.
The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**. This indicates that fundamental changes have taken place in the relationship between the variables during our study period.

@@

Figure \@ref(fig:cumsw) shows the summed hazards of dying in the age interval
40--64 during the different periods. Obviously, but not surprisingly, mortality
decreases over time for both sexes. Women have consistently better survival
than men. This is a well-known phenomenon in Swedish
demographic history [@wil99; @suwi07]. We also observe a
substantial decline in mortality for both sexes, resulting in smaller
absolute differences between social classes. We must, however, distinguish between differences in absolute risks compared to relative risks [@hoetal16]. The compression that is
visible for later decades is partly explained by the decreasing general
mortality and does not necessarily represent a homogenization of relative risks.

By standardizing the levels in each time period for each class, we 
illustrate the relative risks in Figure \@ref(fig:cumswstand). In the rest of 
the paper, we only present these
relative mortality levels for the different classes since they are the main focus in this study.

@@

The changing economy has some consequences for the analysis, particularly when
it comes to the agricultural sector. While farming occupations were common
up until the first half of the 20th century, they have constantly decreased both in numbers as proportionally since then. This means that the middle category has
a different character in the last periods.

The most striking result of the analysis is the different patterns for men
and women. The social pattern of mortality among women is a great deal in
accordance with previous analyses on other regions in northern Sweden and for other periods [@edli11; @edbr12]. The highest social class usually has comparatively low mortality (except for the first period when the statistical power was weak), while
working-class women have high mortality, but almost constantly about the same level as the middle category. However, there is no substantial change in the relative risks between classes except for the periods 1976 onwards when a divide appears between workers and the middle category. A consistent social gradient has developed; the group with unknown social class is distinguished by much higher mortality from the 1976 onwards. This might have to do with how we have defined social position for women before and after 1950. After 1975 it is based on their own occupation; lacking an occupation likely includes women who are more frail when it comes to health or are vulnerable due to social characteristics.

A completely different pattern appears among men, which is the most surprising finding in
our analyses. The groups we expect to have the lowest mortality, if
we assume that access to vital resources and belonging to a high social class
determines survival, instead have by far the highest hazards of dying
during the 19th century and much of the 20th. Workers, on the
other hand, have comparatively low mortality until 1950, at the same
level as the middle class. The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead, we find better survival among workers compared to the elite.

The relative positions change radically towards the end of the 20th century. From the 1980s, there is a clear advantage when it comes to survival for the elite. However, the relative
differences are more compressed compared to the periods until 1950. A reversal of the class pattern in mortality has taken place, and the expected social gradient in mortality has developed. This is also what we usually find in analyses of social health inequalities in present-day
Sweden as well as in most other countries, results that have formed the basis for many of the contemporary epidemiologic and demographic theories. In the regions we are studying,
this is a quite recent phenomenon. Corresponding to what we find
for women, those without a known occupation have much higher mortality than
the others from the late 1970s onwards. The assumption made for women of
unknown social class is probably even more relevant for men, who likely
represent groups in society that are particularly vulnerable.

A further observation can be made from Figure \@ref(fig:cumsw). The gender
difference in death hazards is extremely large for the elite group from
1851--1950, men having about three times higher risks than women in the
same class (basically living in the same households). In the other classes, the difference was much smaller. It was men in high positions who experienced the largest risks.

@@

We now turn to the elderly, those who in most cases have left the workforce.
Corresponding to the previous analysis, we use the value of the cumulative hazards at
duration 25 (age 90), starting from age 65 as a summary measure. Looking at Table \@ref(tab:cumswtabold) it is obvious that the statistical significance is still strong.

@@

The general pattern and development of mortality (not shown here) do not
differ much from the younger age group, of course representing a much
higher mortality level earlier in history. Mortality among women, as
expected, was lower than among men. We find a clear decline in mortality, but
the fastest decline did not set in definitely until sometime after the
Second World War; that is, later than for the age
group 40--64 years. This corresponds to the Swedish development whereby we
find a turning point at this time with better survival among elderly, first for women and a couple of decades later for men.

Concerning the social pattern, the pattern is less clear compared to what
we find for the younger age group in line with the status-levelling hypothesis (see Figure \@ref(fig:cumswold)). For women the results are mixed. We do not find the
same clear advantage for elite women during the period before 1950, but neither
do we find any large disadvantage for working-class elderly women at that
time. In most periods the social differences were small, particularly between the workers and the middle group, and the elite even had somewhat higher mortality during the early 20th century. The social differences continued to be small until the last decades of the
20th century. It is only during the two last periods (after the year 2000) that it is
possible to establish the expected social gradient in mortality.

For men we find indications of the same intriguing change among elderly as
for the adult male population aged 40--64 years, but at a much lower level. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
the 20th century. It is not until the second half of the 20th
century that the survival is better for the highest social class while
workers of retirement age have slightly higher mortality. The difference between the worker and middle group has been small in most periods.

@@

Previous studies have scrutinized different aspects of the social
pattern of mortality in historical northern Sweden for some time, focusing
on different regions, age groups and time periods [@edv92; @edv04; @edli11; @edbr12]. When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economic differences have often, but not always,  been confirmed [@edv92; @edv04],
something that has been found in other Swedish studies as well [@bredro02; @bube01; @mol17].
